Tesseroids: forward modeling gravitational fields in spherical coordinates

### 2015-03-07 manuscript.tex
Replace texttt by emph

The \texttt{src} folder of the source code archive
contains the C files that build the command-line programs
(e.g., \texttt{tessgz.c}).
The \texttt{src/lib} folder contains
the source files that implement the numerical computations.
We will not describe here the implementation of the input/output parsing and
other miscellanea.
Instead, we will focus on the details of the Gauss-Legendre Quadrature
integration of equations~\ref{eq:tesspot}-\ref{eq:tesstensor}
and the adaptive discretization of tesseroids.

@@

The source file \texttt{src/lib/glq.c}
contains the code necessary to perform
a Gauss-Legendre Quadrature integration.
The first step in the GLQ is to compute the
locations of the discretization points (i.e., the point masses).
These points are roots of Legendre polynomials.
Precomputed values are available low order polynomials,
typically up to order five.
For flexibility and to compute higher order roots,
we use the multiple root-finder algorithm of
\citet{Barrera-Figueroa2006}.
The additional computational load is minimal
because the root-finder algorithm
must be run only once per program execution.
The root-finder is implemented in functions
\texttt{glq\_nodes} and \texttt{glq\_next\_root}.
The computed roots will be in the range $[-1, 1]$
and must be scaled to the integration limits
(the physical boundaries of the tesseroid)
using function \texttt{glq\_set\_limits} (see equation~\ref{eq:glq_scaling}).

The GLQ weights (equation~\ref{eq:glq_weights})
are computed by function \texttt{glq\_weights}.
Both the computed roots and weights are stored in a data structure
(a C \texttt{struct}) called \texttt{GLQ}.
Function \texttt{glq\_new}
handles memory allocation,
calculates the roots and weights,
and returns the complete \texttt{GLQ} structure.

The numerical integration of the tesseroid gravitational fields
is performed by the functions in module \texttt{src/lib/grav\_tess.c}.
Functions \texttt{tess\_pot}, \texttt{tess\_gx}, \texttt{tess\_gy}, and so on,
compute the gravitational fields of a single tesseroid
on a single computation point.
These functions require three \texttt{GLQ} structures,
each containing the roots and weights
for GLQ integration in the three dimensions.
The roots must be scaled to the
integration limits
$[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$
(see equations~\ref{eq:tesspot}-\ref{eq:tesstensor}).
The integration consists of three loops
that sum the weighted kernel functions
evaluated at each GLQ point mass (the scaled roots).

@@

We will describe the stack-based algorithm
for computing the effect of a single tesseroid
on a single computation point.
The algorithm is implemented in function \texttt{calc\_tess\_model\_adapt}
of the file \texttt{src/lib/grav\_tess.c}.
The stack of tesseroids is represented by
the \texttt{stack} variable,
an array of \texttt{TESSEROID} structures.
We must define a maximum size for the stack to allocate memory for it.
Defining a maximum size allows us to
avoid an infinite loop
in case the computation point is on
(or sufficiently close to) the surface of the tesseroid.
We use the integer \texttt{stktop}
to keep track of the index of
the last element in the stack (the top of the stack).

The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the input tesseroid.
The initialization is done by copying the tesseroid into the stack
and setting \texttt{stktop} to zero (the first element).
Once the stack is initialized, the steps of the algorithm are:

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \texttt{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ between
        the geometric center of the tesseroid and
        the computation point using equation~\ref{eq:distance}.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \texttt{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\texttt{stktop} is equal to -1).
\end{enumerate}

>>>>

The \emph{src} folder of the source code archive
contains the C files that build the command-line programs
(e.g., \emph{tessgz.c}).
The \emph{src/lib} folder contains
the source files that implement the numerical computations.
We will not describe here the implementation of the input/output parsing and
other miscellanea.
Instead, we will focus on the details of the Gauss-Legendre Quadrature
integration of equations~\ref{eq:tesspot}-\ref{eq:tesstensor}
and the adaptive discretization of tesseroids.

@@

The source file \emph{src/lib/glq.c}
contains the code necessary to perform
a Gauss-Legendre Quadrature integration.
The first step in the GLQ is to compute the
locations of the discretization points (i.e., the point masses).
These points are roots of Legendre polynomials.
Precomputed values are available low order polynomials,
typically up to order five.
For flexibility and to compute higher order roots,
we use the multiple root-finder algorithm of
\citet{Barrera-Figueroa2006}.
The additional computational load is minimal
because the root-finder algorithm
must be run only once per program execution.
The root-finder is implemented in functions
\emph{glq\_nodes} and \emph{glq\_next\_root}.
The computed roots will be in the range $[-1, 1]$
and must be scaled to the integration limits
(the physical boundaries of the tesseroid)
using function \emph{glq\_set\_limits} (see equation~\ref{eq:glq_scaling}).

The GLQ weights (equation~\ref{eq:glq_weights})
are computed by function \emph{glq\_weights}.
Both the computed roots and weights are stored in a data structure
(a C \emph{struct}) called \emph{GLQ}.
Function \emph{glq\_new}
handles memory allocation,
calculates the roots and weights,
and returns the complete \emph{GLQ} structure.

The numerical integration of the tesseroid gravitational fields
is performed by the functions in module \emph{src/lib/grav\_tess.c}.
Functions \emph{tess\_pot}, \emph{tess\_gx}, \emph{tess\_gy}, and so on,
compute the gravitational fields of a single tesseroid
on a single computation point.
These functions require three \emph{GLQ} structures,
each containing the roots and weights
for GLQ integration in the three dimensions.
The roots must be scaled to the
integration limits
$[\lambda_1, \lambda_2], [\phi_1, \phi_2], [r_1, r_2]$
(see equations~\ref{eq:tesspot}-\ref{eq:tesstensor}).
The integration consists of three loops
that sum the weighted kernel functions
evaluated at each GLQ point mass (the scaled roots).

@@

We will describe the stack-based algorithm
for computing the effect of a single tesseroid
on a single computation point.
The algorithm is implemented in function \emph{calc\_tess\_model\_adapt}
of the file \emph{src/lib/grav\_tess.c}.
The stack of tesseroids is represented by
the \emph{stack} variable,
an array of \emph{TESSEROID} structures.
We must define a maximum size for the stack to allocate memory for it.
Defining a maximum size allows us to
avoid an infinite loop
in case the computation point is on
(or sufficiently close to) the surface of the tesseroid.
We use the integer \emph{stktop}
to keep track of the index of
the last element in the stack (the top of the stack).

The algorithm starts by creating an empty stack of tesseroids.
Then, the stack is initialized with the input tesseroid.
The initialization is done by copying the tesseroid into the stack
and setting \emph{stktop} to zero (the first element).
Once the stack is initialized, the steps of the algorithm are:

\begin{enumerate}
    \item ``Pop'' the stack (i.e., take the last tesseroid from it).
        This will cause \emph{stktop} to be reduced by one.
        This tesseroid is the one that will be evaluated in the following
        steps.
    \item Compute the distance $d$ between
        the geometric center of the tesseroid and
        the computation point using equation~\ref{eq:distance}.
    \item Compute the dimensions of the tesseroid $L_\lambda$, $L_\phi$,
        and $L_r$ using equations~\ref{eq:sizelon}-\ref{eq:sizer}.
    \item Check the condition in equation~\ref{eq:condition} for each
        dimension of the tesseroid.
    \item If none of the dimensions fail the above condition,
        compute the gravitational effect of the tesseroid using the
        Gauss-Legendre Quadrature
        (equations~\ref{eq:tesspot}-\ref{eq:tesstensor} and~\ref{eq:glq3d}).
        We use a GLQ order of two for all three dimensions
        ($2 \times 2 \times 2 = 8$ point masses)
        by default.
        This value can be changed using
        a command-line argument of the modeling programs.
    \item If any of the dimensions fail the condition:
    \begin{enumerate}
        \item Divide the tesseroid in half along each dimension that failed
             the condition.
        \item Check if the number of new elements plus \emph{stktop}
             is smaller than the maximum stack size.
             If it is not (i.e., the stack is full),
             warn the user of a ``stack overflow''
             and compute the effect of the larger tesseroid (step 5).
        \item If the stack is not full, place the smaller tesseroids into
             the stack.
    \end{enumerate}
    \item Repeat the above steps until the stack is empty
        (\emph{stktop} is equal to -1).
\end{enumerate}
